---
title: 'W203 Lab 3, Part 1: Reducing Crime'
author:
- Stephen Holtz
- Jon Mease
- Hong Yang
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
```

## Introduction and Research Question
As members of Berkeley Analytica, a political consultancy,
we seek to inform political operatives on what policy decisions
could be most useful once they take office. We also seek to 
help inform their campaigns so they can offer voters a sincere, 
intellectually honest, and meaningful vision for how a candidate 
or party could change society. 

In this project, we are applying a cross section of data from 
C. Cornwell and W. Trumball (1994), "Estimating the Economic Model
of Crime with Panel Data," Review of Economics and Statistics. 
The primary objective of this project is to inform policy makers of the 
value of laws and funding decisions viewed through the lens of reducing
crime. Specific policy questions include, providing funding for more 
police officers; setting guidance or requirements for sentencing of criminals;
implementing policies that improve the distribution of minorities across 
neighborhoods; and implementing policies that keep young males occupied and
out of trouble. 
The secondary objective of this project will be to identify other factors
that affect crime that policy makers could account for in formulating a 
strategy to reduce crime. 

## Initial Data Loading and Cleaning
First we will load and examine the data set.
```{r}
crime_raw <- read.csv("../data/crime_v2.csv")
#crime_raw <- read.csv("crime_v2.csv")
```

We note that the last 6 rows of the dataset are `NA` in all columns except `prbconv`. According to the code book `prbconv` is a numeric variable representing the probability of conviction, but it has been loaded as a factor due to the presence of a backtick string character present in row 97.

```{r}
tail(crime_raw[,1:6], 10)
```

We remove last 6 rows from `crime_raw` and create a new data frame, `crime`, and convert `prbconv` from a factor into a numeric column.
```{r}
crime <- crime_raw[1:(nrow(crime_raw)-6),]
crime$prbconv <- as.numeric(levels(crime$prbconv)[crime$prbconv])
```

Next we examine the probability variables `prbarr` (The 'probability' of arrest), `prbconv` (The 'probability' of conviction), `prbpris` (The probability of prison sentence.).

```{r}
summary(crime[,c('prbarr', 'prbconv', 'prbpris')])
```

Here we see that all probability values are non-negative, but `prbarr` and `prbconv` each have values that are greater than one and therefore not valid probabilities. However, the code book states that "the probability of arrest is proxied by the ratio of arrests to offenses" and "the probability of conviction is proxied by the ratio of convictions to arrests". By these calculations, it is plausible that these probability proxy variables will have values larger than one, so we do not omit these observations.

Next, we note an unreasonably anomalous value for the service industry wage, `wser`, in row 84 for county 185.

```{r}
plot(crime$wser, main = 'Service industry weekly wage')
```

This extreme value (2177.0681) is 860% of the median (253.2) and 556% of the second largest wage in the dataset (391.3081).  Since the other remaining properties for this observation are all in reasonable ranges, with respect to the other observations in the sample, we will replace the anomalous value with `NA` rather than remove the entire observation.

```{r}
crime$wser[84] = NA
```

There is a single unreasonably low value of the `density` variable, which we can identify by plotting density on a log scale.

```{r}
plot(log(crime$density), main = 'Population density (people per square mile)')
```

The 79th observation has a linear scale `density` value of 0.0000203422 people per square mile. The entire state of North Caroline is 53,819 square miles, and if the entire state had this population density there would be only 1.09 persons in the entire state!  Therefore a county that is a small fraction of the size of the state that had this density would have less than one person living in it. As above, the remaining variables for this observation all have reasonable values so we once again replace the errant value by `NA` rather than omit the observation.

```{r}
crime$density[79] = NA
```

## The Model Building Process

To investigate our research question the outcome variable of this analysis will be based on the `crmrte` variable.

### Outcome variable: `crmrte`

Summarizing `crmrte` we see that the mean is a bit larger than the median indicating a moderate positive skew.

```{r}
summary(crime$crmrte)
```

The histogram of `crmrte` confirms the skew and also shows that the distribution of `crmrte` is unimodal.

```{r}
hist(crime$crmrte, main = "Histogram of crime rate")
```

We note that the skew is almost entirely eliminated by taking the natural log of `crmrte`.

```{r}
hist(log(crime$crmrte), main = "Histogram of the log of crime rate")
```

We will delay the decision of whether to use to log of `crmrte` as our outcome variable until we examine the bivariate relationships between crime rate and our explanatory variables of interest.

### Explanatory variables
Because we are working for a political campaign we want to investigate parameters that will inform policy decisions once the political party is in power, as well as the platform that the campaign will use to get elected. 

#### Population density: `density`
We start with `density`, the number of persons per square mile. Here the median is almost 1.5 times larger than the median, indicating a significant positive skew.

```{r}
summary(crime$density)
```

The histogram of `density` confirms the large positive skew and shows that it has a unimodal distribution peaked at around 0.5-1.0 persons per square mile. 

```{r}
hist(crime$density, breaks = 15,
     main = "Histogram of population density")
```

In a scatter plot of `density` and `crmrte` we note a positive relationship between an increase in population density and an increase in crime rate. It appears, however, that the rate of increase in crime rate associated with an increase in density diminishes somewhat as the density increases.

```{r}
plot(crime$density, crime$crmrte,
     main = 'Population density and crime rate')
```

The scatter plot of `log(density)` and `log(crmrte)` shows a much more linear relationship, and so we will use the natural log of `density` as an explanatory variable, and the natural log of `crmrte` and our outcome variable.

```{r}
r2_text <- function(model) {
  return(sprintf('Slope: %.3f\n     R²: %.2f',
                 model$coefficients[[2]], summary(model)$r.squared))
}

model.density <- lm(log(crime$crmrte) ~ log(crime$density))
plot(log(crime$density), log(crime$crmrte),
     main = 'Population density and crime rate')
abline(model.density)
legend('bottomright', legend=r2_text(model.density))
```
Based on the slope of the best-fit line we see that a 1% increase in population density is associated with a ~0.48% increase in crimes per person. Based on the R² statistic we see that population density alone predicts 47% of the variation in crime rate across North Carolina.

#### Probability of arrest, conviction, and prison sentence: `prbarr`, `prbconv`, and `prbpris`

We believe that probability of arrest is worth investigating as it will inform policy decisions and the platform on whether or not more resources should be put towards increasing the likelihood that a person who commits a crime will be arrested.

The histogram of the `prbarr` variable shows that the distribution is unimodal with a slight positive skew.

```{r}
hist(crime$prbarr, breaks = 10,
     main = "Histogram of probability of arrest")
```

In a scatter plot of `prbarr` and `log(crmrte)` we note a negative relationship where an increase in the probability of arrest is associated with a decrease in crime rate.

```{r}
model.prbarr <- lm(log(crime$crmrte) ~ crime$prbarr)
plot(crime$prbarr, log(crime$crmrte),
     main = 'Probability of arrest and crime rate')
abline(model.prbarr)
legend('topright', legend=r2_text(model.prbarr))
```

Based on the slope of the best-fit line we see that an increase of 0.01 units in probability of arrest (one percentage point) is associated with a 1.88% decrease in crime rate. Based on the R² statistic we see that probability of arrest alone predicts 22% of the variation in the crime rate.

In a scatter plot of `prbconv` and `crmrte` we note a negative relationship where an increase in the probability of conviction (given arrest) is associated with a decrease in crime rate.

```{r}
model.prbconv <- lm(log(crime$crmrte) ~ crime$prbconv)
plot(crime$prbconv, log(crime$crmrte),
     main = 'Probability of conviction and crime rate')
abline(model.prbconv)
legend('topright', legend=r2_text(model.prbconv))
```

Based on the slope of the best-fit line we see that an increase of 0.01 units in probability of conviction (one percentage point) is associated with a 0.69% decrease in crime rate. Based on the R² statistic we see that probability of conviction alone predicts 20% of the variation in the crime rate.

Finally, we look at a scatter plot of `prbpris` and `crmrte`. In this case the R² statistic indicates that the probability of a prison sentence (given conviction) only accounts for 0.04% of the variation in crime rate, and so we will omit this variable from our future multivariate models.

```{r}
model.prbpris <- lm(log(crime$crmrte) ~ crime$prbpris)
plot(crime$prbpris, log(crime$crmrte),
     main = 'Probability of prison and crime rate')
abline(model.prbpris)
legend('topright', legend=r2_text(model.prbpris))
```

#### Police officers per capita: `polpc`

A natural instinct of policy makers working to reduce crime may be to increase the number of police officers, and so it is important for them understand the existing association between police officers per capita (`polpc`) and the crime rate.  Here again we apply a log transform, as it once more improves the linearity of the relationship with our outcome variable.

```{r}
model.polpc <- lm(log(crime$crmrte) ~ log(crime$polpc))
plot(log(crime$polpc), log(crime$crmrte),
     main = 'Police officers per capita and crime rate')
abline(model.polpc)
legend('bottomright', legend=r2_text(model.polpc))
```

Based on the slope and R² statistic of the best-fit line we see that a 1% increase in police per capita is actually associated with an *increase* in crime rate of ~0.41%. Based on the R² statistic we see that police officers per capita predict 8% of the variation in crime rate.

#### Percent young male and percent minority: `pctymle` and `pctmin80`

Furthermore, the percentages of young males and minorities in a population need to be examined as those tend to receive ongoing political focus, and the campaign should be aware of the actual relationships between those variables and crime rates.  Motivated by similar observations to those discussed above, we apply a log transform to both of these variables.

```{r}
model.pctmin80 <- lm(log(crime$crmrte) ~ log(crime$pctmin80))
plot(log(crime$pctmin80), log(crime$crmrte),
     main = 'Percent minority and crime rate')
abline(model.pctmin80)
legend('topleft', legend=r2_text(model.pctmin80))
```

```{r}
model.pctymle <- lm(log(crime$crmrte) ~ log(crime$pctymle))
plot(log(crime$pctymle), log(crime$crmrte),
     main = 'Percent young male and crime rate')
abline(model.pctymle)
legend('bottomright', legend=r2_text(model.pctymle))
```

Based on the slope of the best-fit line we see that a 1% increase in the percentage minority population is associated with an increase in crime rate of ~0.16%. Similarly, a 1% increase in the percentage young male population is associated with an increase in crime rate of ~0.86%.

For policy makers, a positive correlation between minorities and crime could be useful evidence to support policies that would lead to greater integration of cultural groups into every neighborhood. Similarly, a positive relationship between young males and crime could be used to justify funding for programs that help young males stay occupied and away from criminal elements.

## Regression Models

### Model 1

Our first model includes the following explanatory variables of key interest to the political campaign: 

 1. The probability of arrest (`prbarr`)
 2. The probability of conviction given arrest (`prbconv`)
 3. Natural log of police per capita (`polpc`)
 
These variables were selected as they are factors that a political operative can potentially influence, once they are in power.

The probability of prison (`prbpris`) and average sentence length (`avgsen`) variables are omitted purposely because they do not provide any meaningful predictive association with the crime rate. This observation is valuable for policy-makers as it could permit better decision making with respect to allocation of funds.

```{r}
(Model1 <- lm(log(crmrte) ~ prbarr + prbconv + log(polpc), data=crime))
sprintf('R²: %.3f', summary(Model1)$r.squared)
```

The model coefficients imply that, all else being equal, an increase in probability of arrest (`prbarr`) of 0.01 (one percentage point) is associated with ~2.34% decrease in the crime rate. Additionally, all else being equal, an increase in the probability of conviction given arrest (`prbconv`) of 0.01 (one percentage point) is associated with a ~0.74% decrease in the crime rate. Surprisingly, it is also found that a 1% increase in police officers per capita is associated with a ~0.6% *increase* in crime rate. The model's R² value indicates that 60.7% of the variation in crime rate can be explained by these three factors alone.

### Model 2

The second model includes the following terms:

 1. The probability of arrest (`prbarr`)
 2. The probability of conviction given arrest (`prbconv`)
 3. Natural log of police per capita (`polpc`)
 4. Natural log of percent minority in 1980 (`pctmin80`)
 5. Natural log of population density (`density`)

The log of density term was added to Model 2 to improve accuracy because the exploratory data analysis revealed that there is a positive correlation between density and crime rate. This correlation has been noted by other researchers (Geoffrey West in "Scale: The Universal Laws of Growth, Innovation, Sustainability, and the Pace of Life in Organisms, Cities, Economies, and Companies"). This is a useful variable as policy makers could encourage development and housing policies that lead to lower population density such as improving transit options to suburban areas.

The log of the percent minority population was added to improve accuracy because, again, exploratory data analysis revealed that there is a positive correlation between and crime rate and this factor.


```{r}
(Model2 = lm(log(crmrte) ~ prbarr + prbconv + log(polpc) +
               log(pctmin80) + log(density), data = crime))
 
sprintf('R²: %.3f', summary(Model2)$r.squared)

```

The `prbarr`, `prbconv`, and `polpc` variables retain their same direction of association as in Model 1, although the magnitude of the associations are reduced somewhat. In addition, we see that a 1% increase in the percent minority population is associated with a ~0.23% increase in the crime rate, and that a 1% increase in the population density is associated with a ~0.23% increase in crime rate.

The model's R² value indicates that 84.8% of the variation in crime rate can be explained by these 5 factors.

### Model 3

Model 3 was constructed using every variable with the exception of the "west", "central", and "urban" variables. We believed that these variables are already correlated with density and they do not have any intrinsic value in informing policy. 

```{r}
(Model3 = lm(log(crmrte) ~  prbarr + prbconv + log(polpc) +
               log(pctmin80) + log(density) + prbpris +
               log(pctymle) + avgsen + wser + county +  taxpc + wcon +
               wtuc + wtrd + wfir + wmfg + wfed + wsta + wloc + mix,
             data = crime))

sprintf('R²: %.3f', summary(Model3)$r.squared)
```

By including all the remaining variables, the r-squared value of Model 3 improved by ~3% percentage points over Model 2. This compares with a ~24% percentage point improvement between Model 2 and Model 1. The analysts at Berkeley Analytica conclude that the improved fit of Model 3 is not justified by the additional data. In short, Model 2 is more valuable as Model 3 may be approaching over-fitting.

### Classical Linear Model Assumptions
Before evaluating the statistical significance our models, it is essential to understand the validity of the Classical Linear Model Assumptions in this case.

#### CLM 1 - Linear in parameters

The linearity observed above between the transformed key relationship variables and the transformed crime rate gives us reasonable confidence that the true relationship between these variables is approximately linear. The team at Berkeley Analytica was certain to transform any explanatory variable selected for a model such that the model would be linear in parameters. Therefore, we can conclude that this assumption is valid.

#### CLM 2 - Random sampling

The researchers and our instructors have provided a dataset for which we are not able to evaluate the assumption of random sampling fully. With what we know from the assignment, we can say that the data is from North Carolina and it includes data from 91 of 100 counties in the state. It is diffucult to assess the validity of the random sampling assumption without further information on why 9 of the counties were omitted.
Further, some thought is required as to the validity of the timeframe over which the crime data was collected. It is our understanding that the crime data was selected from the 1980's, so to evaluate whether the conclusions drawn from this dataset are valid into the future one needs to know why that timeframe was selected. 
Finally, in order to draw conclusions about the causes of crime and to inform policy decisions, the decision to sample only from the state of North Carolina should be highlighted. This data and the conclusions should be viewed through the context of the limited timeframe and the limited geographic scale inherent in the data.

#### CLM 3 - No Multicollinearity

Our exploratory analysis found no perfect collinearity between the variables of key interest.

#### CLM 4 - Zero conditional mean
Given the residuals vs. fitted plots showing there is no clear relationship between the fitted values and the residuals, we can conclude that the zero conditional mean assumption is valid. To the left-hand side of the residuals vs fitted plots graph, there is a deviation of the fit away from zero. However, this is driven by three data points that are all positive. 

```{r}
plot(Model2, which=1)
```

#### CLM 5 - Homoscedasticity

The residual vs. fitted value plot shows the residuals are centered on 0 with relatively constant variance (homoskedasticity) across the range of fitted values, except few values at the higher end.  The Scale-Location plot shows that the average of the square root of the standardized residuals remains nearly constant across the range of fitted values. The only exception is at the far left of the plot, where there are only two observations. Even though this provides strong evidence for homoscedasticity, we will apply the more conservative homoscedasticity robust standard error calculations in our assessment of statistical significence of the model coefficients.

```{r}
plot(Model2, which=3)
```

#### CLM 6 - Normality of error terms
Using histogram or Q-Q plots of the residuals to observe the normality.  The histogram of residuals show that the residuals are distributed approximately normal with mean 0 (with the exception of a few outlying points with larger residuals). 

```{r}
plot(Model2, which=2)
```

```{r}
# normality of standard residuals
stdr = rstandard(Model2)
hist(stdr, main="Standard residuals", breaks = seq(-3.75, 3.75, 0.5), freq=FALSE) 
curve(dnorm(x, mean=0, sd=sd(stdr)), col="red", lwd=2, add=TRUE) 
```

### Models Table

```{r, results = "asis"}

se.Model1 = sqrt(diag(vcovHC(Model1)))
se.Model2 = sqrt(diag(vcovHC(Model2)))
se.Model3 = sqrt(diag(vcovHC(Model3)))

stargazer(Model1, Model2, Model3, type = "latex", 
          title = "Linear Models Predicting Crime Rate",
          omit.stat=c('f', 'adj.rsq'),
          star.cutoffs = c(0.05, 0.01, 0.001),
          se = list(se.Model1, se.Model2, se.Model3),
          omit = c('wcon', 'wtuc', 'wtrd', 'wfir', 'wmfg'),
          #omit.table.layout = "n",
          header=FALSE) # Omit more output related to errors
```

## Omitted Variables Discussion
In order for omitted variables to be causing an over-estimation of the coefficients in our model, two things must be true:

 1. The omitted variables must be correlated with the dependent variable
 2. The omitted variable must be correlated with one or more other independent variables.

Possible candidates for omitted variables might include such varibles or concepts as:

 1. Education or literacy levels, which could be correlated with wages. Presumably, greater education gives individuals more access to the benefits of society and with that, more to lose from antisocial behavior. This can be measured through high school graduation and college graduation rates as well as years of education surveys. If education and literacy rates were included in the model, it is expected that they would be negatively correlated with crime rates, and that the coefficient for density would be pushed higher as a result.
 2. Cost of living, which may be correlated with population density and density. When the cost of living is high, some may feel pressured to undertake criminal endeavors to supplement their incomes. This could be measured using government data on cost of living. Cost of living would be expected to be positively correlated with crime rates, and if it were included in the model, it would be expected to push the density coefficient lower.
 3. Trust in law enforcement and the criminal justice system, which may be correlated with probability of arrest. When the public believes in law enforcement, they are more likely to provide information leading to arrests. This could be measured through surveys as well as the number of tips that are provided for each reported crime. If trust in law enforcement were included in the model, it would be expected to be negatively correlated with crime rates. It is expected that the coefficient for probabilty of arrest would become less negative and that the coefficient for probability conviction would less negative.
 4. Equality of economic opportunity, which may be correlated with wages, percent minority, and percent young male. It has been suggested that when people have access to economic opportunity through legal means, they do not feel the need to fall into criminal activities, and the more egalitarian that opportunity, the fewer people turn to crime. This could be measured by measuring the correlation between parent's income levels and children's income levels. It is believed that equality in economic opportunity would be negatively correlated with crime rates. If this variable were included in the model, it is expected that the percent minority coefficient would be less positive.
 5. Disparity of wealth and/or income, which may be correlated with percent minority, percent  young male, wages, and equality of economic opportunity above. Similar to access to economic opportunity, equality in wealth and/or income may lead to lower crime rates. This could be measured through income tax returns, home value assessments, and vehicle registrations, assuming those three data points provide a reasonable estimate of net worth and income. Disparity in wealth and or income would have a positive correlation with crime rates if it were measured and included in the model. Including this variable in the model would cause the percent minority coefficient to be less positive and the density coefficient to be less positive.
 6. Quality of management of police departments is another possible omitted variable that could correlate to both police per capita and crime rate. Police departments that are well managed would be more effective in reducing crime rates. Quality of management could be measured by identifying the policing tactics that are effective and determining which of these tactics are employed and the degree to which they are employed. Training and competency of the police force could also be measured. The effect of including quality of police management in the model would be to lower the police per capita coefficient as improving quality of management would be negatively correlated with the crime rate.

The results from the polpc (Police per capita) variable is counter-intuitive at first glance. However, on further examination, it becomes clear that both crime rate and police per capita increase with increasing density, however, crime rate increases faster with density than police per capita. The net result is that there is a positive correlation between police and crime. This does not, however, indicate that increasing police on the job will increase the crime rate. 

## Conclusion

The data provided for this study can not be confirmed to meet the random sampling assumptions, and therefore policy analysts should be careful in the application of the conclusions of this study. In particular, policies under consideration thirty years after the data was collected should be checked for validity today. Similarly, when developing policies from this data for  geographies other than North Carolina, the information should be supplemented with local knowledge to correct for the fact that the data was limited to North Carolina.

Based on analysis of the data in this study, the political campaign could argue that to decrease crime rates, policing should be geared towards more and faster arrests. Similarly, the criminal justice system should be geared towards more and faster convictions with shorter prison sentences. Shortening prison sentences would allow for funds to be freed for use in other areas. The deterrent seems to be in arrests and conviction, not as much in being sent to prison, or based on how long a prison sentence is.

Similarly, the political campaign could push for funding programs that keep young males busy and away from criminal elements.

Further studies are needed to understand if value is created for society by having a lower population density and if this value could be
realized by improving transportation options to allow for more suburban developments in favor of urban developments.

Finally, further studies could be performed to see if improving the mix of minorities in neighborhoods helps to reduce crime rates.

Berkeley Analytica believes that a more comprehensive model of the causes of crime could be collected by obtaining and evaluating data on education, cost of living, trust in law enforcement, economic opportunity, economic disparity, and quality of policing. With these variables included in the crime model, Berkeley Analytica believes that further policy recommendations can be developed to help the political operatives win their next election and to govern for the betterment of society.
