---
title: "W203 Lab 3, Part 1: Reducing Crime"
author:
- Stephen Holtz
- Jon Mease
- Hong Yang

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(stargazer)
```

## Introduction and Research Question
As members of Berkeley Analytica, a political consultancy,
we seek to inform political operatives on what policy decisions
could be most useful once they take office. We also seek to 
help inform their campaigns so they can offer voters a sincere, 
intellectually honest, and meaningful vision for how a candidate 
or party could change society. 

In this project, we are applying a cross section of data from 
C. Cornwell and W. Trumball (1994), "Estimating the Economic Model
of Crime with Panel Data," Review of Economics and Statistics. 
The primary objective of this project is to inform policy makers of the 
value of laws and funding decisions viewed through the lens of reducing
crime. Specific policy questions include, providing funding for more 
police officers; setting guidance or requirements for sentencing of criminals;
implementing policies that improve the distribution of minorities across 
neighborhoods; and implimenting policies that keep young males occupued and
out of trouble. 
The secondary objective of this project will be to identify other factors
that affect crime that policy makers could account for in formulating a 
strategy to reduce crime. 

## Initial Data Loading and Cleaning
First we will load and examine the data set.
```{r}
crime_raw <- read.csv("../data/crime_v2.csv")
```

We note that the last 6 rows of the dataset are `NA` in all columns except `prbconv`. According to the codebook `prbconv` is a numeric variable representing the probability of conviction, but it has been loaded as a factor due to the presence of a backtick string character present in row 97.

```{r}
tail(crime_raw[,1:6], 10)
```

We remove last 6 rows from `crime_raw` and create a new data frame, `crime`, and convert `prbconv` from a factor into a numeric column.
```{r}
crime <- crime_raw[1:(nrow(crime_raw)-6),]
crime$prbconv <- as.numeric(levels(crime$prbconv)[crime$prbconv])
```

Next we examine the probability variables `prbarr` (The 'probability' of arrest), `prbconv` (The 'probability' of conviction), `prbpris` (The probability of prison sentence.).

```{r}
summary(crime[,c('prbarr', 'prbconv', 'prbpris')])
```
Here we see that all probability values are non-negative, but `prbarr` and `prbconv` each have values that are greater than one and therefore not valid probabilities. However, the codebook states that "the probability of arrest is proxied by the ratio of arrests to offenses" and "the probability of conviction is proxied by the ratio of convictions to arrests". By these caclulations, it is plausible that these probability proxy variables will have values larger than one, so we do not omin these observations.

Next, we identify an unreasonably anomalous value for the service industry wage, `wser`, in row 84 for county 185.

```{r}
plot(crime$wser, main = 'Service industry weekly wage')
```

This extreme value (2177.0681) is 860% of the median (253.2) and 556% of the second largest wage in the dataset (391.3081).  Since the other remaining properties for this observation are all in reasonable ranges, with respect to the other observations in the sample, we will replace the anomolous value with `NA` rather than remove the entire observation.

```{r}
crime$wser[84] = NA
```


There is a single unreasonably low value of the `density` variable, which we can identify by ploting density on a log scale.

```{r}
plot(log(crime$density), main = 'Population density (people per square mile)')
```

The 79th observation has a linear scale `density` value of 0.0000203422 people per square mile. The entire state of North Caroline is 53,819 square miles, and if the entire state had this population density there would be only 1.09 persons in the entire state!  Therefore a county that is a small fraction of the size of the state that had this density would have less than one person living in it. As above, the remaining variables for this observation all have reasonable values so we once again replace the errant value by `NA` rather than omit the observation.

```{r}
crime$density[79] = NA
```
## The Model Building Process

To investigate our research question the outcome variable of this anlysis will be based on the `crmrte` variable.

### Outcome variable: `crmrte`

Summarizing `crmrte` we see that the mean is a bit larger than the median indicating a moderate positive skew.
```{r}
summary(crime$crmrte)
```

The histogram of `crmrte` confirms the skew and also shows that the distribution of `crmrte` is unimodal.

```{r}
hist(crime$crmrte)
```

We note that the skew is almost entirely eliminated by taking the natural log of `crmrte`.

```{r}
hist(log(crime$crmrte))
```

Based on this, we define our outcome variable as the natrual log of `crmrte`. This means that the models we build in this report will describe percentage (not absolute) changes to the crime rate.


### Explanatory variables
Because we are working for a political campaign we want to investigate parameters that will inform policy decisions once the political party is in power, as well as the platform that the campaign will use to get elected. 

#### Population density: `density`
We start `density`, the number of persons per square mile. Here the median is almost 1.5 times larger than the median, indicating a significant positive skew.

```{r}
summary(crime$density)
```

The histogram of `density` confirms the large positive skew and shows that it has a unimodal distribution peaked at around 0.5-1.0 persons per square mile. 

```{r}
hist(crime$density, breaks = 15)
```

 We will use the natural log of `density` as our explanatory variable as it is significantly less skewed than `density` itself.

```{r}
hist(log(crime$density), breaks = 15)
```

In a scatter plot with our outcome variable we note a positive relationship between percentage change in population density and percentage change in crime rate.

```{r}
r2_text <- function(model) {
  return(sprintf('Slope: %.2f\n     R²: %.2f', model$coefficients[[2]], summary(model)$r.squared))
}

model.density <- lm(log(crime$crmrte) ~ log(crime$density))
plot(log(crime$density), log(crime$crmrte), main = 'Population density and crime rate')
abline(model.density)
legend('bottomright', legend=r2_text(model.density))
```

Based on the slope and R² statistic of the best-fit line we see that a 1% increase in population density is associated with a ~0.49% increase in the crime rate, and that population density by itself can predict 47% of the variation in crime rate across North Carolina.

#### Probability of arrest, conviction, and prison sentence: `prbarr`, `prbconv`, and `prbpris`

We believe that probability of arrest is worth investigating as it will inform policy decisions and the platform on whether or not more resources should be put towards increasing the likelihood that a person who commits a crime will be arrested.

The histogram of the `prbarr` variable shows that the distribution is unimodal with a slight positive skew.
```{r}
hist(crime$prbarr, breaks = 10)
```

Taking the log transform removes the skew altogether, and so we will use the natural log of `prbarr` as an explanotory variable.
```{r}
hist(log(crime$prbarr), breaks = 10)
```

In a scatter plot with our outcome variable we note a negative relationship between percentage change in probability of arrest and percentage change in crime rate.

```{r}
model.prbarr <- lm(log(crime$crmrte) ~ log(crime$prbarr))
plot(log(crime$prbarr), log(crime$crmrte), main = 'Probability of arrest and crime rate')
abline(model.prbarr)
legend('bottomright', legend=r2_text(model.prbarr))
```
Based on the slope and R² statistic of the best-fit line we see that a 1% increase in probability of arrest is associated with a ~0.59% decrease in the crime rate, and that probability of arrest by itself can predict 19% of the variation in crime rate across the state.

Based on similar observations, we also apply a natural log transformation to the probabiliy of conviction (`prbconv`) and probability of prison sentence (`prbpris`) variables.

#### Police officers per capita: `polpc`

A natural instinct of policy makers working to reduce crime may be to increase the number of police officers, and so it is important for them understand the existing associate between police officers per capita (`polpc`) and the crime rate.  Here again we apply a log transform, as it improves the linearity of the relationship with our outcome variable.

```{r}
model.polpc <- lm(log(crime$crmrte) ~ log(crime$polpc))
plot(log(crime$polpc), log(crime$crmrte), main = 'Police officers per capita and crime rate')
abline(model.polpc)
legend('bottomright', legend=r2_text(model.polpc))
```

Based on the slope and R² statistic of the best-fit line we see that a 1% increase in police per capita is associated with a ~0.42% increase in the crime rate, and that this association predicts 8% of the variation in crime rate.

#### Percent young male and percent minority: `pctymle` and `pctmin80`

Furthermore, the percentages of young males and minorities in a population need to be examined as those tend to receive ongoing political focus, and the campaign should be aware of the actual relationships between those variables and crime rates.  For the same reasons as discussed above, we apply a log transform to both of these variables.

```{r}
model.pctmin80 <- lm(log(crime$crmrte) ~ log(crime$pctmin80))
plot(log(crime$pctmin80), log(crime$crmrte), main = 'Percent minority and crime rate')
abline(model.pctmin80)
legend('bottomright', legend=r2_text(model.pctmin80))
```

```{r}
model.pctymle <- lm(log(crime$crmrte) ~ log(crime$pctymle))
plot(log(crime$pctymle), log(crime$crmrte), main = 'Percent young male and crime rate')
abline(model.pctymle)
legend('bottomright', legend=r2_text(model.pctymle))
```

Based on the slope and R² statistic of the best-fit line we see that a 1% increase in percentage minority is associated with a ~0.23% increase in crime rate. Similarly, a 1% increase in percentage young male is associated with a ~0.86% increase in crime rate.

For policy makers, a positive correlation between minorities and crime could be useful evidence to support policies that would lead to greater integration of cultural groups into every neighborhood. Similarly, a positive relationship between young males and crime could be used to justify funding for programs that help young males stay occupied and away from criminal elements.

## Regression Models

### Model 1
Model with only the explanatory variables of key interest.

prbarr(because lower chance of arrest can lead to more crime)
prbconv and prbpris (for similar reason as prbarr)
polpc has strong positive correlation with prbarr, pctmin80 is included for the reason of political campaign.

```{r}
scatterplotMatrix(~log(crmrte)+log(prbarr), data = crime)
```

```{r}
scatterplotMatrix(~log(crmrte)+log(polpc), data = crime)
```

```{r}
(Model1 <- lm(log(crmrte) ~ log(prbarr) + log(prbconv) + log(prbpris) + log(avgsen) + log(polpc), data=crime))
summary(Model1)
```

```{r}
plot(Model1$fitted.values, Model1$residuals)
```

# Model 1 includes the explanatory variables of key interest to the political 
# campaign: 
# 1. Probability of Arrest, transformed by taking the log
# 2. Probability of Conviction
# 3. Probability of Prison
# 4. Police Per Capita, transformed by taking the log
# These variables were selected as they are factors that a political operative
# can influence, once they are in power. The average sentence length was 
# omitted purposely because it did not have a meaningful correlation with crime
# rates. This observation is valuable for policy-makers as it could permit
# better decision making with respect to allocation of funds.

### Model 2
adding density, wfed and wser to the key explanatory variables.
Density and urban are problematic, due to multicollinearity, correlation = .82
Wages - one type goes up the other type tends to go up, they can tend to split the effect among themselves. There is considerable amount of multicollinearity among the different wages.
Also density and wages have some multicollinearity, which make sense since jobs in the city tend to have higher wages.
Residuals apprear to be centered at 0, also we don’t see any patterns in the residuals

```{r}
(Model2 = lm(log(crmrte) ~ log(prbarr) + log(prbconv) + log(prbpris) + log(polpc) + log(pctmin80) + log(density) + wfed + wser, data = crime))
summary(Model2)$r.square
summary(Model2)
```

```{r}
plot(Model2$fitted.values, Model2$residuals)
```

# Model 2 includes the following terms:
# 1. Log of probability of arrest
# 2. Probability of conviction
# 3. Log of Police Per Capita
# 4. Percent Minority
# 5. Log of density
# 6. Average wage of federal employees
# 7. Average wage of service industry employees

# The log of density term was added to Model 2 to improve accuracy
# because the exploratory data analysis revealed that there is a
# positive correlation between density and crime rate. This correlation
# has been noted by other researchers (Geoffrey West in "Scale: The Universal 
# Laws of Growth, Innovation, Sustainability, and the Pace of Life in 
# Organisms, Cities, Economies, and Companies"). This is a useful variable as 
# policy makers could encourage development and housing policies that lead 
# to lower population density such as improving transit options to suburban 
# areas. 

# The average wage of federal employees was included as it was expected to
# add a degree of further accuracy to the model, but as most of the wage terms
# were colinear with density, most of the other wage terms were omitted purposely.

# The average wage of the service industry was included because among the
# wage variables it was unique for being negatively correlated with crime
# rates. Policy makers could view this as evidence for the value of increasing
# the minimum wage in the hopes of reducing crime.

```{r}
(Model2.1 = lm(log(crmrte) ~ log(prbarr) + prbconv + prbpris + log(polpc) + pctmin80 + urban, data = crime))
summary(Model2.1)$r.square
```
# Model2.1 was created to confirm if 'urban' was a better 
# variable to use than 'density' as the two are colinear.
# Model2.1 is less accurate than Model2, so we will continue with 
# Model2.

### Model 3

```{r}
(Model3 = lm(log(crmrte) ~ log(prbarr) + prbconv + prbpris + log(polpc) + pctmin80 + log(density) + wser + county + avgsen + taxpc + wcon + wtuc + wtrd + wfir + wmfg + wfed + wsta + wloc + mix + pctymle, data = crime))
summary(Model3)$r.square
```

```{r}
plot(Model3$fitted.values, Model3$residuals)
```

# Model 3 was constructed using every variable with the exception of the "west", "central", and "urban"
# variables. The analysts believed that these variables are already correlated with density and 
# they do not have any intrinsic value in informing policy. 
# By including all the remaining variables, the r-squared value of Model 3 improved by 
# 4.39 percentage points over Model 2. This compares with a 12 percentage point improvement between
# Model 2 and Model 1. 
# The analysts at Berkeley Analytica conclude that the improved fit of Model 3 is not justified 
# by the additional data. In short, Model 2 is more valueable as Model 3 may be approaching over-fitting.

### Models Table

```{r}
stargazer(Model1, Model2, Model2.1, Model3, type = "text", 
          report = "vc", # Don't report errors, since we haven't covered them
          title = "Linear Models Predicting Crime Rate",
          keep.stat = c("rsq", "n"),
          omit.table.layout = "n",
          header=FALSE) # Omit more output related to errors
```

### Observations
 - The predictive power of pctmin80 is very stable as additional factors are added to the model. Appears to be orthogonal to other features
 - Predictive power of density is decreased some when prob arrest/conviction are added to the model (There is some colinearity between density and these values)
 - prob arrest and conviction are not very correlated with each other and each independently is associated with a decrease in crime
 - polpc correlates positively with ave sentence.

Maybe policy to shift people toward lower density:

TODO: interpret
TODO: State/Evaluate CLM assumptions.

## Omitted Variables Discussion
Need some discussion about `pctmin80` result. We don't have information on the demographics of those being convicted of crimes, 

And `polpc` influence is counterintuitive. Why is there a positive relationship between police per capita and crime?

## Conclusion
Political platform of pushing for more/faster arrest/convictions with no or shorter prison sentences. Deterrent seems to be in arrest/conviction, not as much in being sent to prison, or based on how long a prison sentence is.

Further studies needed to understand
